High-order dynamic neural correlations reflect naturalistic processing in humans

### 2019-05-07 paper/main.tex
WIP paper; finished methods and results, working on discussion

\subsubsection*{Inter-subject dynamic correlations}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}$.  Suppose that one
has access to a set of multiple observation matrices that reflect the
same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\{ mathbf{X}_1,
mathbf{X}_2, ..., mathbf{X}_P\}$ reflect the $T$ by $F$ observation
matrices for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{i=1}^P
  Z\left(Y_i(t)\right)^\mathrm{T} + Z\left(Y_i,(t)\right)\right)\right),
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_i(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_i$ and each
column of the average observations from all \textit{other}
participants, $\bar{\mathbf{X}}_{ \backslash i}$:
\begin{align}
  \bar{\mathbf{X}}_{ \backslash i} = R\left(\frac{1}{P-1}\sum_{i \in
  \backslash i} Z\left( \mathbf{X}_i \right) \right),
\end{align}
where $ \backslash i$ denotes the set of all participants other than
participant $i$. In this way, the $T$ by $\left( \frac{F^2 - F}{2}
  + F \right)$
matrix $\bar{\mathbf{C}}$ is the time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

To generate synthetic data from a dynamically changing covariance
matrix (Ramping, Fig.~\ref{fig:synthetic_data},  C.),
we generated two random covariance
matrices, $K_{1}$ and $K_{2}$.  We then computed a weighted average covariance matrix
for each of the 300 timepoint, $K(t)$, 
by taking the linearly spaced weights ($w$) of the two random matrices, 
\begin{align}
K(t) = w(t) * K_{1} + (1 - w(t)) *  K_{2},
\end{align}
and for each of the 300 timepoints sampled from a
Gaussian distribution centered on $K(t)$. We also explored
recovery by comparing this ramping synthetic data to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  C.) and using only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
C.).

For the synthetic data containing block structure, we followed the same
process of creating synthetic data sampled from a constant covariance
matrix (see above) but sampled from a new random covariance matrix
after each 60 consecutive timepoints.  We then pieced the blocks together
to create a sythetic dataset with 300 total timepoints but drawn from
5 separate covariance matrices. We then compared recovery using the
total dataset (All, Fig.~\ref{fig:synthetic_data},  D.) or limiting to
the block specific covariance matrix (Block 1-5, Fig.~\ref{fig:synthetic_data},  D.).

(Fig.~\ref{fig:synthetic_data},  D.) ,

(Fig.~\ref{fig:synthetic_data})

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data.pdf}
  \caption{\textbf{Dynamic correlation recovery with synthetic data. } Using synthetic data containing different
    underlying correlational structure,
   we test how well we can recover using two kernels (delta and gaussian,
    width = 10).  We plot recovery using of datasets containing the
    following underlying structure:
    \textbf{a. Constant. b. Random. c. Ramping.  d. Block.}}
  \label{fig:synthetic_data}
\end{figure}

@@

We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:weights}) to each of these 
synthetic datasets, then correlate each recovered
correlation matrix with the ground truth.  We explore how recovery varies
with the kernel and the specific structure of the data
(e.g. slow changes as in the ramping dataset, versus rapid changes as
in the block dataset).

@@

\section*{Results}
\subsubsection*{Synthetic data}
Figure: overall timecourse of recovery, also recovery near
event boundaries.

@@

Separately for each experimental condition, we divided participants
into two groups. For the zeroth order, we computed the mean factor
activity for each group.  For all subsequent orders up to the tenth
order, we computed the mean approximated dynamic ISFC of factor
activity for each group (see \textit{Materials and methods}), and
combined in a weighted mixutre with all previous orders
(i.e. cross-validation for the second
order contained a weighted-mixture of zeroth, first, and second order \textbf{c. Optimized weights.}.)  For
each order, we correlated the group 1 activity patterns with group 2
activity patterns.  We then subdivided the group 1 to obtain an
optimal weighting parameter for each order’s correlation matrix using
the same cross validation method. We used the optimal weighting
parameters to obtain a weighted-mixture (see \textit{Materials and methods}) of each order’s correlation
matrix. Using these correlations, we labeled the group 1 timepoints
using the group 2 timepoints with which they were most highly
correlated; we then computed the proportion of correctly labeled group
1 timepoints. (We also performed the symmetric analysis whereby we
labeled the group 2 timepoints using the group 1 timepoints as a
template.) We repeated this procedure 100 times (randomly re-assigning
participants to the two groups each time) to obtain a distribution of
decoding accuracies for each experimental condition. (There were 272
timepoints for paragraph condition, 300 timepoints for intact and word
conditions, and 400 timepoints for rest condition,  so chance
performance on this decoding test is was $\frac{1}{272}$,
$\frac{1}{300}$, and $\frac{1}{400}$ respectively.

@@

>>>>

\subsubsection*{Inter-subject dynamic correlations}
Equation~\ref{eqn:timecorr} provides a means of taking a single
observation matrix, $\mathbf{X}$ and estimating the dynamic
correlations from moment to moment, $\mathbf{Y}$.  Suppose that one
has access to a set of multiple observation matrices that reflect the
same phenomenon.  For example, one might collect neuroimaging data
from several experimental participants, as each participant performs
the same task (or sequence of tasks).  Let $\mathbf{X}_1$,
$\mathbf{X}_2$, ..., $\mathbf{X}_P$ reflect the $T$ by $F$ observation
matrices for each of $P$ participants in an experiment.  We can use
\textit{inter-subject functional connectivity}~\citep[ISFC;
][]{SimoEtal16} to compute the degree of stimulus-driven correlations
reflected in the multi-participant dataset at a given timepoint $t$
using:
\begin{align}
\bar{\mathbf{C}}(t) = M\left(R\left(\frac{1}{2P} \sum_{i=1}^P
  Z\left(Y_i(t)\right)^\mathrm{T} + Z\left(Y_i,(t)\right)\right)\right),
\end{align}
where $M$ extracts and vectorizes the diagonal and upper triangle of a symmetric
matrix, $Z$ is the Fisher $z$-transformation~\citep{Zar10}:
\begin{align}
Z(r) = \frac{\log(1+r) - \log(1-r)}{2}
\end{align}
$R$ is the inverse of $Z$:
\begin{align}
R(z) = \frac{\exp(2z - 1)}{\exp(2z + 1)},
\end{align}
and $\mathbf{Y}_i(t)$ denotes the correlation matrix
(Eqn.~\ref{eqn:corr}) between each column of $\mathbf{X}_i$ and each
column of the average observations from all \textit{other}
participants, $\bar{\mathbf{X}}_{ \backslash i}$:
\begin{align}
  \bar{\mathbf{X}}_{ \backslash i} = R\left(\frac{1}{P-1}\sum_{i \in
  \backslash i} Z\left( \mathbf{X}_i \right) \right),
\end{align}
where $ \backslash i$ denotes the set of all participants other than
participant $\i$. In this way, the $T$ by $\left( \frac{F^2 - F}{2}
  + F \right)$
matrix $\bar{\mathbf{C}}$ is the time-varying extension of the ISFC
approach developed by \cite{SimoEtal16}.

@@

To generate synthetic data from a dynamically changing covariance
matrix (Fig.~\ref{fig:synthetic_data},  C.),
we generated two random covariance
matrices, $K_{1}$ and $K_{2}$.  We then computed a weighted average covariance matrix
for each of the 300 timepoint, $K(t)$, 
by taking the linearly spaced weights ($w$) of the two random matrices, 
\begin{align}
K(t) = w(t) * K_{1} + (1 - w(t)) *  K_{2}, \\\label{eqn:ramping}
\end{align}
and for each of the 300 timepoints sampled from a
Gaussian distribution centered on $K(t)$.

Lastly, for the synthetic data containing block structure (Fig.~\ref{fig:synthetic_data},  D.), we followed the same
process of creating synthetic data sampled from a constant covariance
matrix (see above) but sampled from a new random covariance matrix
after 60 consecutive timepoints.  We then pieced the blocks together
to create a sythetic dataset with 300 total timepoints but drawn from
5 separate covariance matrices.

@@

We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:weights}) to each of these 
synthetic datasets, then correlate each recovered
correlation matrix with the ground truth.  We explore how recovery varies
with the kernel and the specific structure of the data. For the
ramping synthetic dataset (Fig.~\ref{fig:synthetic_data},  C.)  and for the
block synthetic dataset (Fig.~\ref{fig:synthetic_data},  D.)  we made further
comparisons of the timecorr recovered correlation matrices. We
compared the ramping recovered correlation matrices to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  C.) and to only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
C.) from Equation~\ref{eqn:ramping}. We also compared the block recovered correlation matrices in to
the block specific covariance matrix (Block 1-5,
Fig.~\ref{fig:synthetic_data},  D.).

@@

\section*{Results}
\subsubsection*{Synthetic data}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/synthetic_data.pdf}
  \caption{\textbf{Dynamic correlation recovery with synthetic data. } Using synthetic data containing different
    underlying correlational structure,
   we test how well we can recover dynamic correlation matrices using two kernels (delta and gaussian,
    width = 10) when compared to groud truth.  We plot recovery using of datasets containing the
    following underlying structure:
    \textbf{a. Constant. b. Random. c. Ramping.  d. Block.}}
  \label{fig:synthetic_data}
\end{figure}

To assess the performance of dynamic correlation recovery using
timecorr, we varied width the kernel and the specific structure of the data. We applied timecorr, using delta and gaussian kernels
Fig.~\ref{fig:weights}) to each of the following  
synthetic datasets: constant, random, ramping, and block.  We then correlated each recovered
correlation matrix with the ground truth.

For the constant synthetic dataset, a gaussian kernel (width=10)
outperformed the delta kernel (Fig.~\ref{fig:synthetic_data},  A.).  This is in contrast with the random
synthetic dataset, for which the delta kernel best captures the
rapidly changing structure (Fig.~\ref{fig:synthetic_data},  B.). For
the ramping synthetic dataset, the slow changing strucutre within the
data is best
captured by the gaussian kernel and the best recovery occurs in the
middle (Ramping, Fig.~\ref{fig:synthetic_data},
C.). In addition to comparing the timecorr recovered correlation
matrices to the ground truth, we
further compared the ramping recovered correlation matrices to only the first random covariance matrix $K_{1}$
(First, Fig.~\ref{fig:synthetic_data},  C.) and to only the last
random covariance matrix $K_{2}$ (Last, Fig.~\ref{fig:synthetic_data},
C.), both of which perform best at the beginning and end respectively.

Similary for the block sythetic dataset, we compared the timecorr
recovered correlation matrices to the ground truth as well as to each
block-specific covariance matrix (Block 1-5,
Fig.~\ref{fig:synthetic_data},  D.).  Although the structure is
changing by block, the gaussian kernel once again outperforms the
delta kernel.  Performance does however drop near even boundaries for
when using the gaussian kernel.

@@

Separately for each experimental condition, we divided participants
into two groups. For the zeroth order, we computed the mean factor
activity for each group.  For all subsequent orders up to the tenth
order, we computed the mean approximated dynamic ISFC of factor
activity for each group (see \textit{Materials and methods}), and
combined in a weighted mixutre with all previous orders
(i.e. cross-validation for the second
order contained a weighted-mixture of zeroth, first, and second order (Fig.~\ref{fig:decoding_level},  C.).  For
each order, we correlated the group 1 activity patterns with group 2
activity patterns.  We then subdivided the group 1 to obtain an
optimal weighting parameter for each order’s correlation matrix using
the same cross validation method. We used the optimal weighting
parameters to obtain a weighted-mixture (see \textit{Materials and methods}) of each order’s correlation
matrix. Using these correlations, we labeled the group 1 timepoints
using the group 2 timepoints with which they were most highly
correlated; we then computed the proportion of correctly labeled group
1 timepoints. (We also performed the symmetric analysis whereby we
labeled the group 2 timepoints using the group 1 timepoints as a
template.) We repeated this procedure 100 times (randomly re-assigning
participants to the two groups each time) to obtain a distribution of
decoding accuracies for each experimental condition. (There were 272
timepoints for paragraph condition, 300 timepoints for intact and word
conditions, and 400 timepoints for rest condition,  so chance
performance on this decoding test is was $\frac{1}{272}$,
$\frac{1}{300}$, and $\frac{1}{400}$ respectively.

We found that, during each of the listening conditions in the
experiment, classifiers that incorporated second-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  A.).  (Incorporating higher-order correlations did
not further improve decoding accuracy.)  By contrast, these
higher-order correlations were not necessary to support (minimally
above chance) decoding during a control rest condition.  This suggests
that the cognitive processing that supported the listening conditions
involved second-order (or higher) network dynamics.

@@

Dynamic interactions between brain structures underlie our thoughts.
We developed a model of neural dynamics that takes in a feature
timeseries and outputs estimated first-order network dynamics (i.e.,
dynamic functional correlations), second-order network dynamics
(reflecting homologous networks that dynamically form and disperse),
and higher-order network dynamics (up to tenth-order dynamic
correlations).  After validating our model using synthetic data, we
applied the approach to an fMRI dataset
~\citep{SimoEtal16} in which participants listened to an audio recording of a story, as well as scrambled versions of the same story (where the scrambling was applied at different temporal scales).  We trained classifiers to take the output of the model and decode the timepoint in the story (or scrambled story) that the participants were listening to. We found that, during each of the listening conditions in the
experiment, classifiers that incorporated up to second-order correlations
yielded consistently higher accuracy than classifiers trained only on
lower-order patterns (Fig.~\ref{fig:decoding_level},  A.).  (Incorporating higher-order correlations did
not further improve decoding accuracy.)  By contrast, these
higher-order correlations were not necessary to support (minimally
above chance) decoding during a control rest condition.  This suggests
that the cognitive processing that supported the listening conditions
involved second-order (or higher) network dynamics.

Although we found decoding accuracy was best for second-order network dynamics for all but rest
  condition, it is unclear if this is a product of the brain or the
  data collection technique .  It could be that the brain is
  second-order or it could be that fMRI can
  only reliably give second-order interactions.  Exploring this method
  with other data collection technique will be important to
  disentangle this question.
